An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-03 implementation.tex
fixing problems

Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

@@

Consider the first \idsl{telemetry1} property. The property-specific 
part of the synthesized monitor, is shown 
in Figure \ref{fig:monitor}\footnote{An additional 600+ lines of property independent 
boilerplate code is generated.}. This function is called for each new event. The function operates on the two arrays holding BDDs, indexed by subformula indexes: \iscala{pre} for the previous state and \iscala{now} for the current state. For each observed event, the function \iscala{evaluate()} computes the \iscala{now} array, evaluating any sub-formula before any formula containing it,
and returns true (property is satisfied in this position of the trace)  iff \idsl{now(0)} is not $\bfalse$.

\begin{figure}
%\centering
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}[language=scala,basicstyle=\small\sffamily]
override def evaluate(): Boolean = {
  // -- 1. rule nodes not below @:
  now(7) = pre(8)
  now(6) = build("toggle")(V("x"))
  now(5) = now(6).biimp(now(7))
  // -- 2. rule calls:
  now(9) = now(5)
  now(2) = now(5)
  // -- 3. rest of rules below @:
  now(8) = now(9).not()
  // -- 4. rest of main formula:
  now(4) = build("telem")(V("x"))
  now(3) = now(4).not()
  now(1) = now(2).not().or(now(3))
  now(0) = now(1).forAll(var_x.quantvar)

  val error = now(0).isZero
  if (error) monitor.recordResult()
  tmp = now; now = pre; pre = tmp
  !error
}
\end{lstlisting}
\end{tabular}
\end{center}
\caption{Monitor evaluation function for telemetry1}
\label{fig:monitor}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=7cm]{figures/ast.pdf}
\caption{Annotated abstract syntax tree for telemetry1}
\label{fig:ast}
\end{figure}

The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:ast}. The figure (generated by \dejavu) points out (blue node 5) the top node for the rule \idsl{closed(x)}, and calls to the rule (purple nodes 2 and 9).
The subformulas are evaluated in four steps, as outlined at the end of Section \ref{LTLruntime}. First rule nodes not below the previous-operator \idsl{@} (7, 6, 5). Then calls of rules (9, 2)
which can now be calculated since the rule bodies have been calculated. Then the rest of the rules below \idsl{@}  (8). Finally the rest of the main formula (4, 3, 1, 0). At composite subformula nodes, BDD operators are applied. For example for subformula  1, the new  value is \iscala{now(2).not().or(now(3))}, 
which is the interpretation of the formula 
\idsl{closed(x) -> !telem(x)} corresponding to the Boolean law:
$p \rightarrow q = \neg p \vee q$.

@@

\begin{figure}
    \centering   
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd1.pdf}
        \caption{toggle(``low'')}
        \label{fig:bdd1}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd2.pdf}
        \caption{toggle(``high'')}
        \label{fig:bdd2}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd3.pdf}
        \caption{closed(x)}
        \label{fig:bdd3}
    \end{subfigure}
    \caption{Selected BDDs from trace evaluation}
    \label{fig:bdds}
\end{figure}

>>>>

Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
%
The ASCII version of the logic here uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

@@

Consider the first \idsl{telemetry1} property. The property-specific 
part of the synthesized monitor, is shown 
in Figure \ref{fig:monitor} (left)\footnote{An additional 600+ lines of property independent 
boilerplate code is generated.}. This function is called for each new event. The function operates on the two arrays holding BDDs, indexed by subformula indexes: \iscala{pre} for the previous state and \iscala{now} for the current state. For each observed event, the function \iscala{evaluate()} computes the \iscala{now} array, evaluating any sub-formula before any formula containing it,
and returns true (property is satisfied in this position of the trace)  iff \idsl{now(0)} is not $\bfalse$.

\begin{figure}[htb]
\centering
\begin{minipage}[b]{.51\textwidth}
\begin{lstlisting}[language=scala,basicstyle=\sffamily]
override def evaluate(): Boolean = {
  // -- 1. rule nodes not below @:
  now(7) = pre(8)
  now(6) = build("toggle")(V("x"))
  now(5) = now(6).biimp(now(7))
  // -- 2. rule calls:
  now(9) = now(5)
  now(2) = now(5)
  // -- 3. rest of rules below @:
  now(8) = now(9).not()
  // -- 4. rest of main formula:
  now(4) = build("telem")(V("x"))
  now(3) = now(4).not()
  now(1) = now(2).not().or(now(3))
  now(0) = now(1).forAll(var_x.quantvar)

  val error = now(0).isZero
  if (error) monitor.recordResult()
  tmp = now; now = pre; pre = tmp
  !error
}
\end{lstlisting}
\end{minipage}
\qquad
\begin{minipage}[b]{.43\textwidth}
\includegraphics[width=7cm]{figures/ast.pdf}
\end{minipage}
\caption{Monitor (left) and subformula enumeration (right) for property telemetry1}
\label{fig:monitor}
\end{figure}

The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:monitor} (right). The figure (generated by \dejavu) points out (blue node 5) the top node for the rule \idsl{closed(x)}, and calls to the rule (purple nodes 2 and 9).
The subformulas are evaluated in four steps, as outlined at the end of Section \ref{LTLruntime}. First rule nodes not below the previous-operator \idsl{@} (7, 6, 5). Then calls of rules (9, 2)
which can now be calculated since the rule bodies have been calculated. Then the rest of the rules below \idsl{@}  (8). Finally the rest of the main formula (4, 3, 1, 0). At composite subformula nodes, BDD operators are applied. For example for subformula  1, the new  value is \iscala{now(2).not().or(now(3))}, 
which is the interpretation of the formula 
\idsl{closed(x) -> !telem(x)} corresponding to the Boolean law:
$p \rightarrow q = \neg p \vee q$.

@@

\begin{figure}
    \centering   
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd1.pdf}
        \caption{toggle(``low'')}
        \label{fig:bdd1}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd2.pdf}
        \caption{toggle(``high'')}
        \label{fig:bdd2}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd3.pdf}
        \caption{closed(x)}
        \label{fig:bdd3}
    \end{subfigure}
    \caption{Selected BDDs from trace evaluation}
    \label{fig:bdds}
\end{figure}
